---
title: "Semantic Topics in Levebee AI Mathematics Feedback"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Semantic Topics in Levebee AI Mathematics Feedback}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(sbert)
```

# Why this tutorial

The `feedback_translations` dataset bundled with **sbert** contains 8,757
AI-generated feedback messages from the Levebee mathematics application,
translated into English from ten source languages. Nobody can read nearly nine
thousand messages, and many repeat verbatim — "Try again." is the most common,
appearing 11 times — so simple frequency lists tell you about templates,
not content. Topic modeling answers the question the raw data cannot: *what
kind of feedback does the system actually give, and in what proportions?*

The answer has to survive scrutiny, so every step here is deterministic:
rerunning this document reproduces every number, bar, and sentence exactly.
The tutorial first lets the data choose the number of topics, builds the
resulting six-topic model, and then inspects it from three deliberately
different angles:

1. **Model quality** — coherence and diversity, so you know which topics to
   trust before interpreting any of them.
2. **Two keyword views per topic** — raw within-topic counts (what the topic
   *says most*) against class-based TF-IDF (what the topic *alone* says).
   These are different lists by construction, and the disagreement between
   them is itself informative.
3. **Representative sentences** — the centroid-nearest messages, which are
   the auditable evidence that a topic label means what it claims.

| Component | Verb | Question it answers |
|---|---|---|
| Topic count | `compare_topics()` | How many topics does the corpus support? |
| Topic model | `topics()`, `fitted()` | What groups exist? |
| Model quality | `summary()`, `coherence()` | Which topics are trustworthy? |
| Topic size | `topic_sizes()` | How big is each topic, by distinct and reused share? |
| Frequent keywords | `terms(sort_by = "beta")` | What does each topic talk about most? |
| Distinctive keywords | `terms()` | What does each topic talk about that others do not? |
| Evidence | `representatives()` | Do real messages support the label? |

Fourteen revision-pinned models are available; this tutorial uses the
default, `all-MiniLM-L6-v2` (the field's standard quality-per-megabyte
English embedder). The menu, with each model's dimension, input limit,
language coverage, and download size:

```{r menu}
models()
```

# The corpus, deduplicated

Embedding the same string twice wastes computation and — more importantly —
lets repeated templates skew the clustering geometry: every copy of a
message like "Try again." acts as another point pulling on the same
centroid. `dedupe()` collapses the corpus to its distinct non-blank messages
(each kind votes once) while keeping the row frequencies, which return as
weights when sizes are reported. To keep this tutorial fast and free of any
download, it works with the first 600 distinct messages — drop the `head()` to
model the whole corpus:

```{r dedupe}
corpus <- head(dedupe(feedback_translations$translation), 600)
```

`model_download()` is the one step that touches the network. Called without
arguments it fetches the package default, `all-MiniLM-L6-v2` — a 6-layer
distilled English model — as its ONNX graph (90.4 MB) and tokenizer into the
local cache, locked to one immutable Hugging Face revision and refused unless
every byte matches the SHA-256 hashes pinned inside the package, so the model
you run today is provably the model you run next year. To use a different
embedder, pass its name — `model_download("bge-small-en-v1.5")`, for example —
and `models()` lists the other thirteen pinned options. From there
`compare_topics()` and `topics()` do the rest: under the hood they encode each
distinct message — tokenized, run through the network, mean-pooled with padding
masked out, and L2-normalized — into a 600 × 384 matrix, numerically identical
to Python `SentenceTransformers` output and bit-identical on every rerun.

```{r download, eval=FALSE}
model_download()
```

# Choosing the number of topics

There is no universally correct number of topics, and **sbert** deliberately
refuses to pick one for you — the right granularity depends on what the analysis
is *for*. Instead of guessing, `compare_topics()` fits one model per candidate
count and reports the numbers that justify a choice: coherence (do a topic's top
terms actually co-occur?), diversity (do topics own their vocabulary or share
it?), and the share of embedding variance `explained`.

```{r sweep-show, eval=FALSE}
sweep <- compare_topics(
  corpus$text,
  n_topics = c(4, 5, 6, 7, 8, 10, 12),
  measure = "npmi",
  n_representatives = 5
)
sweep
```

```{r sweep, include=FALSE}
# The same 600 messages' embeddings ship precomputed with the package
# (data-raw/feedback_embeddings.R), so the sweep needs no download to build this
# vignette; feeding them in is identical to encoding live.
.sample <- readRDS(
  system.file("extdata", "feedback_embeddings.rds", package = "sbert")
)
stopifnot(identical(corpus$text, .sample$text))
sweep <- compare_topics(
  corpus$text,
  n_topics = c(4, 5, 6, 7, 8, 10, 12),
  embeddings = .sample$embeddings,
  measure = "npmi",
  n_representatives = 5
)
```

```{r sweep-print, echo=FALSE}
sweep
```

```{r sweep-plot, fig.height=4}
plot(sweep)
```

Coherence is negative throughout — these are short, template-like messages, so
any two top terms rarely land in the same short string — but it peaks
unmistakably at **six** and falls away on either side, while diversity stays
high and `explained` keeps climbing with the count. Six is the granularity this
corpus actually supports: the point after which more topics stop buying
coherence, and (not by coincidence) about the number of feedback *kinds* an
editorial review can act on. `fitted()` pulls the already-fitted six-topic model
straight out of the sweep — no re-encoding, no second clustering:

```{r model}
topic_model <- fitted(sweep, n_topics = 6)
```

Only `model_download()` and the encoding inside `compare_topics()` are skipped
here — the 600 messages' embeddings ship precomputed — so the vignette builds
with no download. Run these lines yourself and you reproduce every number below
exactly.

```{r model-print}
topic_model
```

## Should you trust these topics?

Interpretation comes *after* evaluation, because a beautifully labeled topic
with poor coherence is a story about noise. NPMI coherence asks whether a
topic's top terms actually co-occur in its messages (+1 = always together,
−1 = never); diversity asks whether topics share their vocabulary or own it.

```{r evaluate}
summary(topic_model)
```

```{r coherence}
coherence(topic_model, measure = "npmi")
```

Carry these numbers into the per-topic panels below: NPMI rewards regular,
repetitive phrasing, so the most template-like topics tend to score highest,
and any topic with visibly lower coherence should be read through its
representative sentences rather than its keywords.

## Size on two scales

The model counts *distinct* messages, but the application sent some messages
many times over, so editorial priority follows the weighted share, not the
distinct share. `topic_sizes()` reports both scales in one call; the gap
between `proportion` and `weighted_share` measures how template-driven each
topic is — a topic whose weighted share far exceeds its distinct share is a
small repertoire of heavily reused messages.

```{r sizes, fig.height=5}
plot(topic_model, type = "sizes")
```

```{r weighted}
topic_sizes(topic_model, weights = corpus$n)
```

# Two keyword views, one topic

`terms(sort_by = "beta")` returns the empirical probability of each word given
the topic — the *generative* view, dominated by whatever the topic says most
often. The class-based TF-IDF scores of `terms()` (its default `sort_by`) are
the *discriminative* view: they promote words this topic uses and others do not,
and demote words that recur across many topics — frequent, but saying little
about any single one. Neither list is "the" keywords; a topic is characterized
by the pair. When the two lists agree — as they do for topic 1 below, whose most
frequent and most distinctive words are the same ("picture", "pictures",
"choose") — the topic has a vocabulary of its own; when they disagree, the counts
list is telling you about the corpus and only the TF-IDF list about the topic.

```{r beta}
head(terms(topic_model, n = 8, sort_by = "beta"), 8)
```

```{r terms}
head(terms(topic_model, n = 8), 8)
```

# The topics at a glance

`plot(topic_model, type = "terms")` shows each topic's distinctive class-based
TF-IDF terms, each bar annotated with its score:

```{r plot-terms, fig.height=6}
plot(topic_model, type = "terms")
```

`type = "fit"` is the whole per-topic report: all three keyword views — raw
within-topic count, class-based TF-IDF, and generative probability (beta) —
alongside the topic's centroid-nearest messages, one row per topic:

```{r plot-fit, fig.width = 8, fig.height = 11}
plot(topic_model, type = "fit", n_terms = 8, n_representatives = 3)
```

With `per_topic = TRUE` each topic becomes its own figure, the documents stacked
beneath the terms — easier to read one topic at a time:

```{r plot-fit-single, fig.height=4.5}
plot(topic_model, type = "fit", per_topic = TRUE, topics = 1)
```

The plot text is clipped to fit; for the full, untruncated evidence behind each
topic, ask for the representatives directly:

```{r representatives}
representatives(topic_model, n = 2)
```

# Where to go from here

The model built here is reusable, not just describable. `predict()` assigns
any new feedback message to these six topics without refitting;
`topic_membership()` replaces the hard assignment with graded probabilities
when a message sits between topics. Multi-sentence messages can be modelled
sentence by sentence instead of whole: `segment = "sentence"` on the fitting
verbs splits every message first and fits the topics on the sentences, and
`compare_topics()` takes several levels at once so the unit is chosen from a
table and a plot, exactly like the count. The fitted sentence model stores each
sentence with its parent message, so `topic_gamma()` returns every message's
topic mixture with no further encoding and `topic_sizes(by = "document")`
counts messages rather than sentences:

```{r segments, eval=FALSE}
comparison <- compare_topics(
  corpus$text,
  n_topics = c(4, 6, 8),
  segment = c("document", "sentence")
)
plot(comparison)
plot(comparison, type = "fit", n_topics = 6, n_terms = 8, n_representatives = 3)

sentence_model <- fitted(comparison, n_topics = 6, segment = "sentence")
topic_gamma(sentence_model)
topic_sizes(sentence_model, by = "document", weights = corpus$n)
```

When you explore many models on the same corpus — the sweep above, or several
term settings — `topic_corpus()` segments, embeds, and tokenizes the messages
once and every fit reuses that work, so the exploration costs one corpus pass
rather than one per model. For multilingual work on the *source* messages (the `feedback` column
spans ten languages), swap one argument —
`topics(corpus$text, n_topics = 6, model = "paraphrase-multilingual-MiniLM-L12-v2")` —
and the rest of this document runs unchanged.
