---
title: "Topic Modeling COVID-19 Research Abstracts"
subtitle: "A semantic review of the pandemic education literature, 2020-2024"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  dpi = 120,
  out.width = "100%"
)
library(sbert)
```

<style>
body { color: #25364a; font-size: 15px; }
h1, h2, h3, h4 { color: #153e5c; }
.kpi-grid { display: grid; grid-template-columns: repeat(5, minmax(120px, 1fr)); gap: 12px; margin: 18px 0 24px; }
.kpi { background: #f3f7fa; border-top: 4px solid #247ba0; border-radius: 6px; padding: 13px 15px; }
.kpi-value { color: #153e5c; font-size: 24px; font-weight: 700; }
.kpi-label { color: #526579; font-size: 11px; letter-spacing: .04em; text-transform: uppercase; }
.caveat { background: #fff7dd; border-left: 6px solid #d99000; border-radius: 5px; padding: 15px 18px; margin: 18px 0 24px; }
.method-note { background: #edf8f5; border-left: 6px solid #238b78; border-radius: 5px; padding: 15px 18px; margin: 18px 0 24px; }
.topic-block { border: 1px solid #d6e0e8; border-radius: 7px; padding: 14px 18px; margin: 16px 0; background: #fff; }
.topic-block h3 { margin-top: 0; }
.topic-meta { color: #5d6d7e; margin-bottom: 12px; }
.term-label { background: #f4f7f9; border-left: 4px solid #247ba0; padding: 10px 13px; margin: 10px 0; }
details { border-top: 1px solid #e3e8ed; padding: 9px 0; }
summary { color: #153e5c; cursor: pointer; font-weight: 600; }
.paper-meta { color: #5d6d7e; font-size: 13px; margin: 6px 0; }
.abstract { white-space: normal; line-height: 1.45; }
table { font-size: 13px; }
th { background: #247ba0 !important; color: #fff !important; }
</style>

The bundled `covid` dataset holds abstracts of COVID-19 research on education,
children, schools, and society. This review embeds every abstract with a
Sentence-BERT model, groups them into semantic topics, and reads the result —
including how the mix of topics shifted across the pandemic years. Everything
is deterministic: rerunning reproduces every number and figure.

# Building the model

Drop the records indexed without an abstract, then clean the source text before
encoding. `clean_corpus()` repairs junk characters, strips any list and
reference numbering, and — with `min_content` — drops whole low-content rows
such as citation lists. It carries the other columns along, so `Year` stays
aligned. Sentence-BERT tolerates a little noise, so the goal is fixing broken
text and removing non-content rows, not scrubbing every token. On a few messy
examples the effect is easiest to see:

```{r clean-demo, echo=FALSE}
messy <- c(
  "1. The programme raised attainment in 2020, see section 3.2.1.",
  "(3) OJ No L 297, 24.11.1979, p. 1.",
  "Peer&nbsp;support <b>improved</b> outcomes [12] across schools.",
  "Full report at https://example.org/study.pdf and doi:10.1/x."
)
knitr::kable(
  data.frame(Before = messy, After = clean_corpus(messy)),
  caption = "clean_corpus() with defaults: list and reference numbering, HTML tags, bracketed citations, URLs and DOIs removed and characters repaired, while real words are kept."
)
```

The citation-only row above scores far below the prose rows on alphabetic
density, so `min_content = 0.5` drops it entirely rather than embedding it. The
generic cleaners deliberately stop short of domain-specific reference forms; for
those, `remove` takes custom patterns — `clean_corpus(text, remove = "OJ No
L?\\s*\\d+")` strips European Union Official Journal identifiers, for example.
The same clean text then feeds the encoder, the topic terms, and the
representative abstracts.

```{r corpus}
covid_content <- covid[covid$Abstract != "[No abstract available]", ]
covid_content <- clean_corpus(covid_content, column = "Abstract", min_content = 0.5)
```

`encode()` turns each abstract into a 384-dimensional vector with the pinned
default model, `all-MiniLM-L6-v2`. Abstracts longer than the model's 256-token
window are truncated to their opening; for full-length documents a long-context
model such as `nomic-embed-text-v1.5` (8,192 tokens) is a drop-in replacement.

```{r embed}
install_runtime()
model_download()

embeddings <- encode(covid_content$Abstract, batch_size = 32)
```

Every abstract shares a scientific boilerplate — *covid*, *pandemic*, *study*,
*results* — and the label of *every* topic would otherwise repeat it. Those
words carry no signal about what separates one topic from another, so they are
excluded up front; `numbers = "remove"` drops bare years and counts (*2020*,
*19*) for the same reason. Domain stop words are the standard way to keep
topic labels informative.

```{r stopwords}
covid_stops <- stop_words(add = c(
  "covid", "coronavirus", "sars", "cov", "pandemic", "disease",
  "study", "studies", "result", "results", "conclusion", "conclusions",
  "background", "method", "methods", "objective", "aim"
))
```

There is no correct topic count. `select_topics()` fits one model per candidate
and reports the numbers that justify a choice:

```{r sweep}
sweep <- select_topics(
  covid_content$Abstract,
  n_topics = c(4, 6, 8, 10, 12),
  embeddings = embeddings,
  measure = "npmi",
  stop_words = covid_stops,
  numbers = "remove"
)
sweep
```

```{r sweep-plot, fig.height=4.5}
plot(sweep)
```

Read the count from the table rather than by habit: coherence rises, peaks, and
falls as topics multiply, while `explained` keeps climbing regardless. The
count with the highest coherence is the granularity this corpus best supports —
the point after which splitting topics stops buying coherence — so the model is
fitted there.

```{r model}
best_n <- sweep$n_topics[which.max(sweep$coherence)]
topic_model <- fitted(sweep, n_topics = best_n)
```

# Corpus at a glance

```{r audit-kpis, results='asis', echo=FALSE}
kpi <- function(value, label) {
  sprintf(
    '<div class="kpi"><div class="kpi-value">%s</div><div class="kpi-label">%s</div></div>',
    value, label
  )
}
cat(
  '<div class="kpi-grid">',
  kpi(format(nrow(covid), big.mark = ","), "Records"),
  kpi(format(nrow(covid_content), big.mark = ","), "Modeled abstracts"),
  kpi(format(length(unique(covid_content$Abstract)), big.mark = ","), "Distinct"),
  kpi("2020-2024", "Years"),
  kpi(nrow(topic_model$topics), "Topics"),
  '</div>'
)
```

<div class="caveat">
<strong>Read the topics as a lens, not a taxonomy.</strong> The count was chosen
because coherence peaks there, not because it is uniquely correct. Labels are
the three most distinctive class-based TF-IDF terms — with corpus boilerplate
and bare numbers excluded — not validated names. A bibliographic export is never
perfectly on-theme, so one small cluster collects genuinely off-topic papers
(veterinary and materials research) — the model isolating them rather than
contaminating the education topics.
</div>

# The topics

```{r overview, echo=FALSE}
overview <- topic_model$topics[, c("topic", "label", "n_documents", "proportion")]
overview$proportion <- sprintf("%.1f%%", 100 * overview$proportion)
names(overview) <- c("Topic", "Distinctive terms", "Abstracts", "Share")
knitr::kable(overview, row.names = FALSE)
```

`plot(type = "fit")` lays out every topic at once — the three keyword views
(raw within-topic count, class-based TF-IDF, and generative probability) beside
each topic's representative abstracts, one row per topic:

```{r plot-fit, fig.width = 11, fig.height = 15}
plot(topic_model, type = "fit", n_terms = 8, n_representatives = 8)
```

The cards below give the same evidence topic by topic, with the full abstracts
in collapsible panels — the auditable proof that a label means what it claims.

```{r topic-blocks, results='asis', echo=FALSE}
html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  gsub(">", "&gt;", x, fixed = TRUE)
}
compact <- function(x, n = 700L) {
  x <- trimws(gsub("[[:space:]]+", " ", x))
  ifelse(nchar(x) > n, paste0(substr(x, 1L, n - 1L), "..."), x)
}

reps <- representatives(topic_model, n = 3)
reps$year <- covid_content$Year[match(reps$text, covid_content$Abstract)]
palette <- topic_palette(nrow(topic_model$topics))

blocks <- vapply(seq_len(nrow(topic_model$topics)), function(i) {
  tid <- topic_model$topics$topic[[i]]
  label <- topic_model$topics$label[[i]]
  terms_i <- utils::head(
    topic_model$terms$term[topic_model$terms$topic == tid], 8L
  )
  tr <- reps[reps$topic == tid, , drop = FALSE]
  tr <- tr[order(tr$rank), , drop = FALSE]
  papers <- vapply(seq_len(nrow(tr)), function(j) {
    sprintf(
      paste0(
        '<details%s><summary>Nearest abstract %d - %s (distance %.3f)</summary>',
        '<p class="abstract">%s</p></details>'
      ),
      if (j == 1L) " open" else "",
      j,
      ifelse(is.na(tr$year[[j]]), "year n/a", as.character(tr$year[[j]])),
      tr$distance[[j]],
      html_escape(compact(tr$text[[j]]))
    )
  }, character(1))
  sprintf(
    paste0(
      '<div class="topic-block" style="border-left:6px solid %s">',
      '<h3>Topic %d &mdash; %s</h3>',
      '<div class="topic-meta">%s abstracts &middot; %.1f%% of the corpus</div>',
      '<div class="term-label"><strong>Distinctive terms:</strong> %s</div>%s</div>'
    ),
    palette[[i]],
    tid,
    html_escape(label),
    format(topic_model$topics$n_documents[[i]], big.mark = ","),
    100 * topic_model$topics$proportion[[i]],
    html_escape(paste(terms_i, collapse = ", ")),
    paste(papers, collapse = "\n")
  )
}, character(1))
cat(paste(blocks, collapse = "\n"))
```

# What each paper touches

The document model gives each abstract a single label — its dominant theme. A
long abstract usually spans several topics, though, and unlike the whole-abstract
embedding, its individual sentences are short enough to escape the 256-token
truncation. `segment()` splits a document into sentences, deterministically —
and `max_tokens` guards against the rare sentence that is itself over the
window, re-splitting it at punctuation first. The cap counts words here, a
deterministic offline proxy; 200 words sits comfortably under the 256-token
limit (pass `model =` to count that model's exact tokens instead):

```{r segment-demo}
segments <- segment(covid_content$Abstract, level = "sentence", max_tokens = 200)
head(segments, 3)
```

`topic_gamma()` takes that segmentation directly — assigning every sentence to
its nearest topic and returning each abstract's topic *mixture* (`gamma` sums to
1 within a document). Passing the `segment()` frame keeps the cap and lets a
single encode be reused rather than segmenting again inside the call:

```{r gamma}
gamma <- topic_gamma(topic_model, segments)
head(gamma)
```

Most abstracts touch several topics. One that spreads across the most, with the
share of its sentences on each:

```{r gamma-one, echo=FALSE}
spread <- tapply(gamma$gamma, gamma$document_id, function(x) sum(x > 0.1))
mixed_id <- as.integer(names(which.max(spread)))
mix <- subset(gamma, document_id == mixed_id & gamma > 0.05)
mix <- mix[order(-mix$gamma), ]
knitr::kable(
  data.frame(
    Topic = topic_model$topics$label[match(mix$topic, topic_model$topics$topic)],
    `Share of sentences` = sprintf("%.0f%%", 100 * mix$gamma),
    check.names = FALSE
  ),
  row.names = FALSE
)
```

Aggregating every sentence gives a second view of prevalence — the share of
*sentences* on each topic — next to the document-level share. The two agree on
the big themes but diverge on the secondary ones: a topic that is rarely a
paper's headline can still run through many papers' sentences.

```{r gamma-prevalence, echo=FALSE, fig.height=5}
agg <- tapply(gamma$gamma * gamma$n_segments, gamma$topic, sum)
sent_share <- as.numeric(agg[as.character(topic_model$topics$topic)])
sent_share <- sent_share / sum(sent_share)
mat <- rbind(document = topic_model$topics$proportion, sentence = sent_share)
colnames(mat) <- paste0("T", topic_model$topics$topic)
old_par <- graphics::par(mar = c(4, 4, 3, 11))
graphics::barplot(
  mat, beside = TRUE, border = NA,
  col = c("#247ba0", "#d99000"),
  xlab = "Topic", ylab = "Share",
  main = "Topic prevalence: whole-abstract vs sentence-level"
)
graphics::legend(
  "topright", inset = c(-0.34, 0), xpd = NA, bty = "n", cex = 0.8,
  fill = c("#247ba0", "#d99000"),
  legend = c("document (headline)", "sentence (touched)")
)
graphics::par(old_par)
```

A topic model of whole abstracts reports each paper's *headline*; the sentence
view shows everything it *touches* — and because sentences are never truncated,
it reaches content the 256-token window cuts off.

# Topics across the pandemic

Because every abstract carries a publication year, the model doubles as a lens
on how the research conversation moved. The share of each year's abstracts
assigned to each topic:

```{r temporal, fig.height=6, echo=FALSE}
prevalence <- prop.table(
  table(topic = topic_model$documents$topic, Year = covid_content$Year),
  margin = 2
)
old_par <- graphics::par(mar = c(4, 4, 3, 15))
graphics::barplot(
  prevalence,
  col = palette,
  border = NA,
  xlab = "Year",
  ylab = "Share of abstracts",
  main = "Topic composition of COVID-19 research by year"
)
graphics::legend(
  "topright",
  inset = c(-0.44, 0),
  xpd = NA,
  bty = "n",
  cex = 0.75,
  fill = palette,
  legend = topic_model$topics$label
)
graphics::par(old_par)
```

Each bar is one year; the coloured segments are the topics' shares that year.
Reading left to right shows which strands of the education literature grew and
which faded as the pandemic wore on.

<div class="method-note">
<strong>Reusing the model.</strong> `predict()` assigns new abstracts to these
topics without refitting, and `topic_membership()` gives graded probabilities
when an abstract sits between topics. For long abstracts, `segment()` splits
each into sentences and `blend()` carries the paper's context into every
sentence's embedding, so a sentence that is ambiguous alone still embeds near
its subject.
</div>
