---
title: "Parallel and High-Throughput Topic Modeling"
subtitle: "Running sbert at full speed without changing a single result"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(sbert)
```

<style>
body { color: #25364a; font-size: 15px; }
h1, h2, h3 { color: #153e5c; }
.callout { background: #edf8f5; border-left: 6px solid #238b78; border-radius: 5px; padding: 15px 18px; margin: 20px 0; }
.warn { background: #fff7dd; border-left: 6px solid #d99000; border-radius: 5px; padding: 15px 18px; margin: 20px 0; }
table { font-size: 14px; }
th { background: #247ba0 !important; color: #fff !important; }
code { color: #14514a; }
</style>

A topic model in **sbert** is built in three stages, and each scales
differently:

1. **Encode** — text becomes Sentence-BERT vectors, through the ONNX model.
2. **Tokenize** — text becomes terms, for the topic labels and coherence.
3. **Cluster** — vectors become topics, through deterministic k-means.

The largest speed-ups come from two ideas: **do the corpus-level work once**
(encoding and tokenization depend on the corpus, not on the number of topics),
and **spread that work across cores**. Crucially, neither changes the output —
every result below is byte-identical to a plain single-threaded run.

# Do the expensive work once

Encoding is almost always the most expensive stage, and tokenization is the
next after it once the corpus is large. Both depend only on the documents, so
computing them once and reusing them is the single biggest lever when you fit
more than one model — a topic-count sweep, a comparison of term settings, or any
loop over `n_topics`.

**Pass precomputed embeddings.** Encode once, then hand the matrix to every fit:

```{r embed-once}
model <- load_model("all-MiniLM-L6-v2", threads = 8)
embeddings <- encode(corpus$text, model, batch_size = 64)

# topics()/select_topics() now skip encoding entirely
m <- topics(corpus$text, n_topics = 8, embeddings = embeddings)
```

**Prepare a corpus once, fit many.** `topic_corpus()` goes further: it caches
*both* the embeddings and the tokenization, so a sweep or loop repeats neither.

```{r corpus-once}
prepared <- topic_corpus(corpus$text, embeddings = embeddings)

sweep <- select_topics(prepared, n_topics = c(6, 8, 10, 12, 15))
model8 <- fitted(sweep, n_topics = 8)   # chosen without refitting
```

`select_topics()` already shares one prepared corpus across all candidates
internally, so even `select_topics(corpus$text, ...)` no longer re-tokenizes per
candidate. Passing a prepared `topic_corpus` on top lets you reuse it across
*several* sweeps or manual loops.

# The two parallel knobs

## `threads` — parallel encoding

`load_model(threads = N)` sets the number of ONNX inference threads. Every
`encode()` call that uses the loaded model then runs multithreaded:

```{r threads}
model <- load_model("all-MiniLM-L6-v2", threads = 8)
embeddings <- encode(text, model)     # inference spread over 8 threads
```

## `cores` — parallel tokenization and sweeps

`cores` is available on `topics()`, `select_topics()`, `topic_corpus()`, and
`coherence()`. Above one it forks worker processes to:

- **tokenize** the corpus in parallel (documents split across workers), and
- in `select_topics()`, **fit the independent candidates** in parallel.

```{r cores}
# tokenization runs in parallel; the corpus is prepared once
prepared <- topic_corpus(corpus$text, embeddings = embeddings, cores = 8)

# candidates are fitted across cores, tokenization already cached
sweep <- select_topics(prepared, n_topics = c(6, 8, 10, 12, 15), cores = 8)
```

<div class="warn">
**Platform notes.** `cores` uses `parallel::mclapply`, which forks — available
on Linux and macOS, and **falls back to serial on Windows**. Forking has
overhead, so parallel tokenization only engages for corpora of **2,000
documents or more**; smaller inputs run serially by design. A good default is
`cores = parallel::detectCores() - 1`.
</div>

# The maximum-speed recipe

Putting it together for the common case — one corpus, several models:

```{r recipe}
library(sbert)
install_runtime()

# 1. Multithreaded encoding, once
model <- load_model("all-MiniLM-L6-v2", threads = 8)
embeddings <- encode(corpus$text, model, batch_size = 64)

# 2. Prepare the corpus once (encode + tokenize cached), tokenizing in parallel
prepared <- topic_corpus(corpus$text, embeddings = embeddings, cores = 8)

# 3. Sweep across cores, then pick a model with no refit
sweep <- select_topics(prepared, n_topics = c(6, 8, 10, 12, 15), cores = 8)
model8 <- fitted(sweep, n_topics = 8)
```

For a corpus so large that encoding dominates and some quality can be traded for
throughput, the static `potion-base-8M` model embeds around 10,000 sentences per
second in pure R, with no ONNX Runtime involved:

```{r static}
fast <- load_model("potion-base-8M")   # 31 MB, Model2Vec token lookup
embeddings <- encode(corpus$text, fast)
```

# What to expect

Measured on a 10-core Apple Silicon machine (macOS, Accelerate BLAS, R 4.5)
with precomputed embeddings. Wall-clock carries some machine-load variance;
the speed-ups grow with the number of models fitted and with document length.

| Workload | Baseline | Serial | 8 cores |
|---|---:|---:|---:|
| Sweep, 5 counts, 15k short docs | 163.5 s | 12.2 s (13×) | **6.9 s (24×)** |
| Sweep, 3 counts, 5k long docs (~8k chars) | 321.1 s | 56.0 s (5.7×) | **10.3 s (31×)** |
| Tokenization, 15k short docs | — | 11.7 s | **3.1 s (3.8×)** |
| Tokenization, 5k long docs | — | 52.5 s | **8.6 s (6.1×)** |

The "baseline" column is the pre-reuse path that re-encoded and re-tokenized on
every model. Two patterns stand out: longer documents make parallel tokenization
pay off *more* (6.1× vs 3.8×), and the corpus-reuse win scales with the number
of models — roughly 3.5× at five models and 4× at ten.

Memory improves independently of document length, because clustering works on
the embeddings, not the raw text: at 40 topics, peak memory for a 40,000-document
run dropped from about 2,195 MB to 1,053 MB.

# Reproducibility

<div class="callout">
**`cores` and `threads` change speed, never results.** Tokenization is split by
document and recombined in order, and sweep candidates are independent and
deterministic, so the output is identical for any core or thread count. There is
no random seed anywhere in the topic pipeline.
</div>

The proof runs offline — the same model, fitted single-threaded and across four
cores, is byte-identical:

```{r repro, eval = TRUE}
set.seed(1)
n <- 2500L
text <- paste(
  "document", seq_len(n),
  sample(c("alpha beta gamma", "delta epsilon zeta", "eta theta iota"), n, TRUE)
)
# clustered stand-in embeddings, so no model download is needed
centers <- matrix(rnorm(8L * 64L), 8L, 64L)
embeddings <- centers[sample(8L, n, replace = TRUE), ] +
  matrix(rnorm(n * 64L, sd = 0.6), n, 64L)

serial   <- topics(text, n_topics = 6, embeddings = embeddings, cores = 1)
parallel <- topics(text, n_topics = 6, embeddings = embeddings, cores = 4)

identical(
  serial[c("documents", "topics", "terms", "representatives", "centers")],
  parallel[c("documents", "topics", "terms", "representatives", "centers")]
)
```

<div class="warn">
**One knob is not free: `batch_size`.** Larger batches speed up ONNX encoding,
but batch composition perturbs the embeddings at roughly the 1e-7 level, which
can flip a rare cluster assignment. Pick a `batch_size` and keep it — results
are reproducible for a fixed value, just not necessarily across different ones.
</div>

# Checklist

- **Fitting many models on one corpus?** Use `topic_corpus()` (or precompute
  `embeddings`) so encoding and tokenization happen once. This is where the
  order-of-magnitude wins are.
- **Encoding is the bottleneck?** Raise `load_model(threads = N)` and
  `batch_size`, or switch to the `potion-base-8M` static model.
- **On Linux or macOS with a big corpus (≥ 2,000 docs)?** Add
  `cores = detectCores() - 1` to `topic_corpus()` / `select_topics()`.
- **On Windows, or a small corpus?** `cores` safely no-ops to serial; rely on
  reuse and `threads` instead.
- **Need exact reproducibility?** `cores` and `threads` are safe at any value;
  fix `batch_size` once and leave it.
