---
title: "Semantic Topics in LLM Reasoning Traces"
output: html_document
---

```{r setup, include=FALSE}
library(sbert)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(sbert.download = TRUE)
```

The corpus is the `thinking` column of the LLM results: one reasoning trace
per student profile. `dedupe()` drops blanks, keeps one row per distinct trace,
and records each trace's original frequency in `n`.

```{r data}
traces <- rio::import("For_LLM-results.csv")
corpus <- dedupe(traces$thinking)
nrow(corpus)
```

## Four ways to cut a trace, one comparison

A trace runs to a few thousand words, far past what a sentence encoder reads
before truncating. `compare_topics()` fits every candidate topic count on four
units at once: the whole document, and its sentences, clauses, and phrases.
Segments longer than the model's window are re-split, tiny shards are merged
into a neighbour, and numeral-only fragments are dropped. One table, one plot.

```{r compare, fig.width=11, fig.height=4.5}
comparison <- compare_topics(
  corpus$text,
  n_topics = c(4, 6, 8, 10, 12),
  segment = c("document", "sentence", "clause", "phrase"),
  max_tokens = 200,
  merge_below = 4,
  min_content = 0.5,
  n_representatives = 5,
  cores = 4
)
comparison
plot(comparison)
```

Read the coherence panel first: it asks whether a topic's top terms actually
co-occur in its units. Diversity asks whether the topics share their vocabulary
or own it, and explained variance always rises with more topics, so it is a
sanity check rather than a criterion. The document line is the truncated view;
the three segment lines model the traces in full.

## The same six topics, four ways

`type = "fit"` on the comparison draws the per-topic report for every level's
six-topic model: count, class-based TF-IDF, and beta term bars beside the
units nearest each centroid. One figure per level, same layout, so the
segmentations can be read against each other.

```{r compare-fit, fig.width=10.5, fig.height=14}
plot(comparison, type = "fit", n_topics = 6, n_terms = 6, n_representatives = 4)
```

## The model

`fitted()` takes the chosen count and level straight out of the comparison,
with no refitting.

```{r model}
topic_model <- fitted(comparison, n_topics = 6, segment = "sentence")
topic_model
summary(topic_model)
```

## Size on three scales

Segments per topic, source traces per topic, and traces weighted by how often
each was repeated in the raw file. A trace counts in every topic one of its
sentences lands in, so document proportions need not sum to one.

```{r sizes, fig.height=5}
plot(topic_model, type = "sizes")
topic_sizes(topic_model, by = "document", weights = corpus$n)
```

Because the model was fitted on sentences, each trace's mixture over the topics
is already stored. `topic_gamma()` returns it with no re-encoding.

```{r gamma}
head(topic_gamma(topic_model), 12)
```

## Two keyword views, one topic

`sort_by = "beta"` orders each topic's words by how often the topic uses them,
the generative view. The default class-based TF-IDF score promotes words this
topic uses and the others do not, the discriminative view. A topic is
characterised by the pair.

```{r beta}
terms(topic_model, n = 8, sort_by = "beta")
```

```{r terms}
terms(topic_model, n = 5)
```

## The topics at a glance

```{r plot-terms, fig.height=7}
plot(topic_model, type = "terms")
```

The fit report clips its text. `representatives()` ranks the fitted sentences by
distinctiveness margin and names the trace and position each one came from.

```{r representatives}
representatives(topic_model, n = 2)
```

## Where to go from here

`predict()` assigns new traces sentence by sentence with the same
segmentation, `topic_membership()` grades each sentence across topics, and
`reduce_topics()` merges the six down to fewer while keeping every verb above.
